---
title: "preprocess_data"
author: "Aditya Khan"
date: "2026-03-08"
output: pdf_document
---

```{r}

library(data.table)

```


```{r}

stex <- read.csv("data/stex.csv")
toefl <- read.csv("data/toefl_scores_by_native_language.csv")
setDT(stex)
setDT(toefl)

```


```{r}

fix_lang <- function(x) {
  x <- enc2utf8(x)

  x[x %chin% c("\xc9w\xe9", "Éwé", "Ewe")] <- "Ewe"
  x[x == "Byelorussian"] <- "Belarusian"
  x[x == "Portugese"] <- "Portuguese"
  x[x == "Ukranian"] <- "Ukrainian"
  x[x == "Tigrigna"] <- "Tigrinya"

  x
}

lang_lookup <- data.table(
  lang = c(
    "Ewe","Afrikaans","Albanian","Amharic","Arabic","Armenian","Azerbaijani",
    "Bengali","Bosnian","Bulgarian","Belarusian","Catalan","Chinese","Croatian",
    "Czech","Danish","Dutch","English","Estonian","Finnish","French","Fulani",
    "Georgian","German","Greek","Haitian","Hebrew","Hindi","Hungarian","Icelandic",
    "Igbo","Indonesian","Italian","Japanese","Javanese","Kinyarwanda","Korean",
    "Kurdish","Latvian","Lithuanian","Malay","Malayalam","Mongolian","Monolingual",
    "Nepali","Norwegian","Papiamentu","Pashto","Persian","Polish","Portuguese",
    "Romanian","Romansch","Rundi","Russian","Serbian","Slovak","Slovenian",
    "Somali","Spanish","Swahili","Swedish","Tagalog","Tamazight","Tamil","Thai",
    "Tigrinya","Turkish","Ukrainian","Urdu","Vietnamese","Wolof","Yoruba"
  ),
  glottocode = c(
    "ewee1241","afri1274","alba1267","amha1245","arab1395","nucl1235","azer1255",
    "beng1280","bosn1245","bulg1262","bela1254","stan1289","mand1415","croa1245",
    "czec1258","dani1285","dutc1256","stan1293","esto1258","finn1318","stan1290","fula1264",
    "nucl1302","stan1295","mode1248","hait1244","hebr1245","hind1269","hung1274","icel1247",
    "nucl1417","indo1316","ital1282","nucl1643","java1254","kiny1244","kore1280",
    "kurd1259","latv1249","lith1251","stan1306","mala1464","mong1331",NA_character_,
    "nepa1254","norw1258","papi1253","pash1269","cent2390","poli1260","port1283",
    "roma1327","roma1326","rund1242","russ1263","serb1264","slov1269","slov1268",
    "soma1255","stan1288","swah1253","swed1254","taga1270","cent2194","tami1289","thai1261",
    "tigr1271","nucl1301","ukra1253","urdu1245","viet1252","nucl1347","yoru1245"
  )
)

stex[, `:=`(
  L1_std = fix_lang(L1),
  L2_std = fix_lang(L2)
)]

stex[lang_lookup, L1_code := i.glottocode, on = c(L1_std = "lang")]
stex[lang_lookup, L2_code := i.glottocode, on = c(L2_std = "lang")]

stex[, c("L1_std", "L2_std") := NULL]

```


```{r}

angular_distance <- function(u, v, distance_metric = "angular") {
  if (length(u) != length(v)) {
    stop("Vectors must be of the same length")
  }

  u <- as.numeric(u)
  v <- as.numeric(v)

  if (sqrt(sum(u^2)) == 0) {
    u <- rep(1e-10, length(u))
  }
  if (sqrt(sum(v^2)) == 0) {
    v <- rep(1e-10, length(v))
  }

  norm_u <- u / sqrt(sum(u^2))
  norm_v <- v / sqrt(sum(v^2))

  cos_theta <- sum(norm_u * norm_v)
  cos_theta <- max(-1.0, min(1.0, cos_theta))

  if (distance_metric == "cosine") {
    return(round(1 - cos_theta, 4))
  }

  angular_distance <- 2 * acos(cos_theta)
  angular_distance <- angular_distance / pi

  round(angular_distance, 4)
}

pair_distance <- function(code1, code2, lookup, distance_metric = "angular") {
  if (is.na(code1) || is.na(code2)) return(NA_real_)
  if (!(code1 %in% names(lookup)) || !(code2 %in% names(lookup))) return(NA_real_)

  u <- lookup[[code1]]
  v <- lookup[[code2]]

  keep <- !is.na(u) & !is.na(v)
  if (!any(keep)) return(NA_real_)

  u <- u[keep]
  v <- v[keep]

  if (length(u) == 0 || length(v) == 0) return(NA_real_)

  angular_distance(u, v, distance_metric = distance_metric)
}

pair_distance_vec <- function(code1, code2, lookup, distance_metric = "angular") {
  mapply(
    FUN = function(a, b) pair_distance(a, b, lookup, distance_metric = distance_metric),
    code1, code2,
    USE.NAMES = FALSE
  )
}

```


```{r}
feat_dir <- "data/Data/Data/without_glottolog_with_lineage"

features_dt <- fread(file.path(feat_dir, "features.csv"))
family_dt <- fread(file.path(feat_dir, "family_features.csv"))
geo_dt <- fread(file.path(feat_dir, "geocoord_features.csv"))
script_dt <- fread(file.path(feat_dir, "script_features.csv"))

```

```{r}

coerce_feature_numeric <- function(dt) {
  feat_cols <- setdiff(names(dt), "lang")
  for (j in feat_cols) {
    set(dt, j = j, value = as.numeric(dt[[j]]))
  }
  dt
}

features_dt <- coerce_feature_numeric(features_dt)
family_dt <- coerce_feature_numeric(family_dt)
geo_dt <- coerce_feature_numeric(geo_dt)
script_dt <- coerce_feature_numeric(script_dt)

```


```{r}

all_feat_cols <- setdiff(names(features_dt), "lang")
syn_cols <- grep("^S_", all_feat_cols, value = TRUE)
morph_cols <- grep("^M_", all_feat_cols, value = TRUE)
phon_cols <- grep("^P_", all_feat_cols, value = TRUE)
inv_cols <- grep("^INV_", all_feat_cols, value = TRUE)

cat("features:", length(all_feat_cols), "\n")
cat("syntactic:", length(syn_cols), "\n")
cat("morphological:", length(morph_cols), "\n")
cat("phonological:", length(phon_cols), "\n")
cat("inventory:", length(inv_cols), "\n")

```
```{r}

make_vector_lookup <- function(dt, cols) {
  x <- dt[, c("lang", cols), with = FALSE]
  split_vals <- split(x, x$lang)

  out <- lapply(split_vals, function(z) {
    as.numeric(z[1, ..cols])
  })
  out
}

feat_lookup <- make_vector_lookup(features_dt, all_feat_cols)
syn_lookup <- make_vector_lookup(features_dt, syn_cols)
morph_lookup <- make_vector_lookup(features_dt, morph_cols)
phon_lookup <- make_vector_lookup(features_dt, phon_cols)
inv_lookup <- make_vector_lookup(features_dt, inv_cols)

gen_lookup <- make_vector_lookup(family_dt, setdiff(names(family_dt), "lang"))
geo_lookup <- make_vector_lookup(geo_dt, setdiff(names(geo_dt), "lang"))
script_lookup <- make_vector_lookup(script_dt, setdiff(names(script_dt), "lang"))

```

# Old Preprocessing (This is mistaken/incomplete for stex)

```{r}

stex_pairs <- unique(
  stex[, .(
    code1 = pmin(L1_code, L2_code),
    code2 = pmax(L1_code, L2_code)
  )]
)

stex_pairs[, featural_dist := pair_distance_vec(code1, code2, feat_lookup)]
stex_pairs[, script_dist := pair_distance_vec(code1, code2, script_lookup)]
stex_pairs[, geographic_dist := pair_distance_vec(code1, code2, geo_lookup)]
stex_pairs[, genetic_dist := pair_distance_vec(code1, code2, gen_lookup)]
stex_pairs[, syntactic_dist := pair_distance_vec(code1, code2, syn_lookup)]
stex_pairs[, morphological_dist := pair_distance_vec(code1, code2, morph_lookup)]
stex_pairs[, phonological_dist := pair_distance_vec(code1, code2, phon_lookup)]
stex_pairs[, inventory_dist := pair_distance_vec(code1, code2, inv_lookup)]

stex[, `:=`(
  code1 = pmin(L1_code, L2_code),
  code2 = pmax(L1_code, L2_code)
)]

stex[stex_pairs, on = .(code1, code2), `:=`(
  featural_dist = i.featural_dist,
  script_dist = i.script_dist,
  geographic_dist = i.geographic_dist,
  genetic_dist = i.genetic_dist,
  syntactic_dist = i.syntactic_dist,
  morphological_dist = i.morphological_dist,
  phonological_dist = i.phonological_dist,
  inventory_dist = i.inventory_dist
)]

stex[, c("code1", "code2") := NULL]

```


```{r}

english_code <- "stan1293"

toefl[, L1_code := code]
toefl[, L2_code := english_code]

toefl[, featural_dist := pair_distance_vec(L1_code, L2_code, feat_lookup)]
toefl[, script_dist := pair_distance_vec(L1_code, L2_code, script_lookup)]
toefl[, geographic_dist := pair_distance_vec(L1_code, L2_code, geo_lookup)]
toefl[, genetic_dist := pair_distance_vec(L1_code, L2_code, gen_lookup)]
toefl[, syntactic_dist := pair_distance_vec(L1_code, L2_code, syn_lookup)]
toefl[, morphological_dist := pair_distance_vec(L1_code, L2_code, morph_lookup)]
toefl[, phonological_dist := pair_distance_vec(L1_code, L2_code, phon_lookup)]
toefl[, inventory_dist := pair_distance_vec(L1_code, L2_code, inv_lookup)]

```


```{r}

stex[, .(
  n = .N,
  missing_feat = sum(is.na(featural_dist)),
  missing_script = sum(is.na(script_dist)),
  missing_geo = sum(is.na(geographic_dist)),
  missing_gen = sum(is.na(genetic_dist))
)]

toefl[, .(
  n = .N,
  missing_feat = sum(is.na(featural_dist)),
  missing_script = sum(is.na(script_dist)),
  missing_geo = sum(is.na(geographic_dist)),
  missing_gen = sum(is.na(genetic_dist))
)]

```

```{r}

fwrite(stex, "data/stex_with_uriel_distances.csv")
fwrite(toefl, "data/toefl_with_uriel_distances.csv")

```


TODO:
- Think about what to do with Pashto
- Add stuff from [here](https://databank.worldbank.org/source/world-development-indicators/), maybe IT.NET.USER.ZS, NY.GDP.PCAP.PP.KD, SE.TER.ENRR, SE.XPD.TOTL.GD.ZS, SE.ADT.LITR.ZS, SP.URB.TOTL.IN.ZS, SM.POP.TOTL.ZS
- Add stuff from [here](https://hdr.undp.org/data-center/human-development-index#/indicies/HDI)
- Probably do the above only for stex
- For toefl and stex, download [ASJP](https://asjp.clld.org/) so we can use their distances as a competitor
- I thought about including walks/phoible based distances, but could get away with not, since they are included in uriel


# Extra Preprocessing

```{r}

stex.augmented <- read.csv("data/stex_with_uriel_distances.csv")
setDT(stex.augmented)

```


```{r}

dist_cols <- grep("(_dist$)|(^hyperbolic_dist$)", names(stex.augmented), value = TRUE)

setnames(
  stex.augmented,
  old = dist_cols,
  new = paste0(dist_cols, "_l1l2")
)

setnames(
  stex.augmented,
  old = c("morph", "lex", "new_feat", "new_sounds"),
  new = c("Morph_Score", "Lex_Score", "New_Feature_Score", "New_Sound_Score")
)

```


```{r}

dutch_code <- "dutc1256"

l1_dutch <- unique(
  stex.augmented[, .(L1_code)]
)

l1_dutch[, featural_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, feat_lookup)]
l1_dutch[, script_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, script_lookup)]
l1_dutch[, geographic_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, geo_lookup)]
l1_dutch[, genetic_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, gen_lookup)]
l1_dutch[, syntactic_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, syn_lookup)]
l1_dutch[, morphological_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, morph_lookup)]
l1_dutch[, phonological_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, phon_lookup)]
l1_dutch[, inventory_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, inv_lookup)]

l2_dutch <- unique(
  stex.augmented[, .(L2_code)]
)

l2_dutch[, featural_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, feat_lookup)]
l2_dutch[, script_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, script_lookup)]
l2_dutch[, geographic_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, geo_lookup)]
l2_dutch[, genetic_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, gen_lookup)]
l2_dutch[, syntactic_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, syn_lookup)]
l2_dutch[, morphological_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, morph_lookup)]
l2_dutch[, phonological_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, phon_lookup)]
l2_dutch[, inventory_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, inv_lookup)]

stex.augmented[l1_dutch, on = .(L1_code), `:=`(
  featural_dist_l1dutch = i.featural_dist_l1dutch,
  script_dist_l1dutch = i.script_dist_l1dutch,
  geographic_dist_l1dutch = i.geographic_dist_l1dutch,
  genetic_dist_l1dutch = i.genetic_dist_l1dutch,
  syntactic_dist_l1dutch = i.syntactic_dist_l1dutch,
  morphological_dist_l1dutch = i.morphological_dist_l1dutch,
  phonological_dist_l1dutch = i.phonological_dist_l1dutch,
  inventory_dist_l1dutch = i.inventory_dist_l1dutch
)]

stex.augmented[l2_dutch, on = .(L2_code), `:=`(
  featural_dist_l2dutch = i.featural_dist_l2dutch,
  script_dist_l2dutch = i.script_dist_l2dutch,
  geographic_dist_l2dutch = i.geographic_dist_l2dutch,
  genetic_dist_l2dutch = i.genetic_dist_l2dutch,
  syntactic_dist_l2dutch = i.syntactic_dist_l2dutch,
  morphological_dist_l2dutch = i.morphological_dist_l2dutch,
  phonological_dist_l2dutch = i.phonological_dist_l2dutch,
  inventory_dist_l2dutch = i.inventory_dist_l2dutch
)]

```


```{r}
fwrite(stex.augmented, "data/stex_with_uriel_distances_edit.csv")

```


```{r}

stex.new <- read.csv("data/stex_with_uriel_distances.csv")
setDT(stex.new)

```


```{r}

cols_to_suffix <- c(
  "island_dist",
  "speaker_dist",
  "hyperbolic_dist",
  "l2v_genetic_dist",
  "l2v_geographic_dist",
  "l2v_syntactic_dist",
  "l2v_inventory_dist",
  "l2v_phonological_dist",
  "l2v_featural_dist",
  "asjp_dist"
)

new_names <- paste0(cols_to_suffix, "_l1dutch")
data.table::setnames(stex.new, old = cols_to_suffix, new = new_names)

stex.augmented[, (intersect(names(stex.augmented), new_names)) := NULL]
stex.augmented[, (new_names) := stex.new[, ..new_names]]

```


```{r}

fwrite(stex.augmented, "data/stex_with_uriel_distances.csv")

```

