---
title: "preprocess_data"
author: "Aditya Khan"
date: "2026-03-08"
output: pdf_document
---

```{r}

library(data.table)
library(countrycode)

```


```{r}

stex <- read.csv("data/stex.csv")
toefl <- read.csv("data/toefl_scores_by_native_language.csv")
setDT(stex)
setDT(toefl)

```


```{r}

fix_lang <- function(x) {
  x <- enc2utf8(x)

  x[x %chin% c("\xc9w\xe9", "Éwé", "Ewe")] <- "Ewe"
  x[x == "Byelorussian"] <- "Belarusian"
  x[x == "Portugese"] <- "Portuguese"
  x[x == "Ukranian"] <- "Ukrainian"
  x[x == "Tigrigna"] <- "Tigrinya"

  x
}

lang_lookup <- data.table(
  lang = c(
    "Ewe","Afrikaans","Albanian","Amharic","Arabic","Armenian","Azerbaijani",
    "Bengali","Bosnian","Bulgarian","Belarusian","Catalan","Chinese","Croatian",
    "Czech","Danish","Dutch","English","Estonian","Finnish","French","Fulani",
    "Georgian","German","Greek","Haitian","Hebrew","Hindi","Hungarian","Icelandic",
    "Igbo","Indonesian","Italian","Japanese","Javanese","Kinyarwanda","Korean",
    "Kurdish","Latvian","Lithuanian","Malay","Malayalam","Mongolian","Monolingual",
    "Nepali","Norwegian","Papiamentu","Pashto","Persian","Polish","Portuguese",
    "Romanian","Romansch","Rundi","Russian","Serbian","Slovak","Slovenian",
    "Somali","Spanish","Swahili","Swedish","Tagalog","Tamazight","Tamil","Thai",
    "Tigrinya","Turkish","Ukrainian","Urdu","Vietnamese","Wolof","Yoruba"
  ),
  glottocode = c(
    "ewee1241","afri1274","alba1267","amha1245","arab1395","nucl1235","azer1255",
    "beng1280","bosn1245","bulg1262","bela1254","stan1289","mand1415","croa1245",
    "czec1258","dani1285","dutc1256","stan1293","esto1258","finn1318","stan1290","fula1264",
    "nucl1302","stan1295","mode1248","hait1244","hebr1245","hind1269","hung1274","icel1247",
    "nucl1417","indo1316","ital1282","nucl1643","java1254","kiny1244","kore1280",
    "kurd1259","latv1249","lith1251","stan1306","mala1464","mong1331",NA_character_,
    "nepa1254","norw1258","papi1253","pash1269","cent2390","poli1260","port1283",
    "roma1327","roma1326","rund1242","russ1263","serb1264","slov1269","slov1268",
    "soma1255","stan1288","swah1253","swed1254","taga1270","cent2194","tami1289","thai1261",
    "tigr1271","nucl1301","ukra1253","urdu1245","viet1252","nucl1347","yoru1245"
  )
)

stex[, `:=`(
  L1_std = fix_lang(L1),
  L2_std = fix_lang(L2)
)]

stex[lang_lookup, L1_code := i.glottocode, on = c(L1_std = "lang")]
stex[lang_lookup, L2_code := i.glottocode, on = c(L2_std = "lang")]

stex[, c("L1_std", "L2_std") := NULL]

```


```{r}

angular_distance <- function(u, v, distance_metric = "angular") {
  if (length(u) != length(v)) {
    stop("Vectors must be of the same length")
  }

  u <- as.numeric(u)
  v <- as.numeric(v)

  if (sqrt(sum(u^2)) == 0) {
    u <- rep(1e-10, length(u))
  }
  if (sqrt(sum(v^2)) == 0) {
    v <- rep(1e-10, length(v))
  }

  norm_u <- u / sqrt(sum(u^2))
  norm_v <- v / sqrt(sum(v^2))

  cos_theta <- sum(norm_u * norm_v)
  cos_theta <- max(-1.0, min(1.0, cos_theta))

  if (distance_metric == "cosine") {
    return(round(1 - cos_theta, 4))
  }

  angular_distance <- 2 * acos(cos_theta)
  angular_distance <- angular_distance / pi

  round(angular_distance, 4)
}

pair_distance <- function(code1, code2, lookup, distance_metric = "angular") {
  if (is.na(code1) || is.na(code2)) return(NA_real_)
  if (!(code1 %in% names(lookup)) || !(code2 %in% names(lookup))) return(NA_real_)

  u <- lookup[[code1]]
  v <- lookup[[code2]]

  keep <- !is.na(u) & !is.na(v)
  if (!any(keep)) return(NA_real_)

  u <- u[keep]
  v <- v[keep]

  if (length(u) == 0 || length(v) == 0) return(NA_real_)

  angular_distance(u, v, distance_metric = distance_metric)
}

pair_distance_vec <- function(code1, code2, lookup, distance_metric = "angular") {
  mapply(
    FUN = function(a, b) pair_distance(a, b, lookup, distance_metric = distance_metric),
    code1, code2,
    USE.NAMES = FALSE
  )
}

```


```{r}
feat_dir <- "data/Data/Data/without_glottolog_with_lineage"

features_dt <- fread(file.path(feat_dir, "features.csv"))
family_dt <- fread(file.path(feat_dir, "family_features.csv"))
geo_dt <- fread(file.path(feat_dir, "geocoord_features.csv"))
script_dt <- fread(file.path(feat_dir, "script_features.csv"))

```

```{r}

coerce_feature_numeric <- function(dt) {
  feat_cols <- setdiff(names(dt), "lang")
  for (j in feat_cols) {
    set(dt, j = j, value = as.numeric(dt[[j]]))
  }
  dt
}

features_dt <- coerce_feature_numeric(features_dt)
family_dt <- coerce_feature_numeric(family_dt)
geo_dt <- coerce_feature_numeric(geo_dt)
script_dt <- coerce_feature_numeric(script_dt)

```


```{r}

all_feat_cols <- setdiff(names(features_dt), "lang")
syn_cols <- grep("^S_", all_feat_cols, value = TRUE)
morph_cols <- grep("^M_", all_feat_cols, value = TRUE)
phon_cols <- grep("^P_", all_feat_cols, value = TRUE)
inv_cols <- grep("^INV_", all_feat_cols, value = TRUE)

cat("features:", length(all_feat_cols), "\n")
cat("syntactic:", length(syn_cols), "\n")
cat("morphological:", length(morph_cols), "\n")
cat("phonological:", length(phon_cols), "\n")
cat("inventory:", length(inv_cols), "\n")

```
```{r}

make_vector_lookup <- function(dt, cols) {
  x <- dt[, c("lang", cols), with = FALSE]
  split_vals <- split(x, x$lang)

  out <- lapply(split_vals, function(z) {
    as.numeric(z[1, ..cols])
  })
  out
}

feat_lookup <- make_vector_lookup(features_dt, all_feat_cols)
syn_lookup <- make_vector_lookup(features_dt, syn_cols)
morph_lookup <- make_vector_lookup(features_dt, morph_cols)
phon_lookup <- make_vector_lookup(features_dt, phon_cols)
inv_lookup <- make_vector_lookup(features_dt, inv_cols)

gen_lookup <- make_vector_lookup(family_dt, setdiff(names(family_dt), "lang"))
geo_lookup <- make_vector_lookup(geo_dt, setdiff(names(geo_dt), "lang"))
script_lookup <- make_vector_lookup(script_dt, setdiff(names(script_dt), "lang"))

```

# Old Preprocessing (This is mistaken/incomplete for stex)

```{r}

stex_pairs <- unique(
  stex[, .(
    code1 = pmin(L1_code, L2_code),
    code2 = pmax(L1_code, L2_code)
  )]
)

stex_pairs[, featural_dist := pair_distance_vec(code1, code2, feat_lookup)]
stex_pairs[, script_dist := pair_distance_vec(code1, code2, script_lookup)]
stex_pairs[, geographic_dist := pair_distance_vec(code1, code2, geo_lookup)]
stex_pairs[, genetic_dist := pair_distance_vec(code1, code2, gen_lookup)]
stex_pairs[, syntactic_dist := pair_distance_vec(code1, code2, syn_lookup)]
stex_pairs[, morphological_dist := pair_distance_vec(code1, code2, morph_lookup)]
stex_pairs[, phonological_dist := pair_distance_vec(code1, code2, phon_lookup)]
stex_pairs[, inventory_dist := pair_distance_vec(code1, code2, inv_lookup)]

stex[, `:=`(
  code1 = pmin(L1_code, L2_code),
  code2 = pmax(L1_code, L2_code)
)]

stex[stex_pairs, on = .(code1, code2), `:=`(
  featural_dist = i.featural_dist,
  script_dist = i.script_dist,
  geographic_dist = i.geographic_dist,
  genetic_dist = i.genetic_dist,
  syntactic_dist = i.syntactic_dist,
  morphological_dist = i.morphological_dist,
  phonological_dist = i.phonological_dist,
  inventory_dist = i.inventory_dist
)]

stex[, c("code1", "code2") := NULL]

```


```{r}

english_code <- "stan1293"

toefl[, L1_code := code]
toefl[, L2_code := english_code]

toefl[, featural_dist := pair_distance_vec(L1_code, L2_code, feat_lookup)]
toefl[, script_dist := pair_distance_vec(L1_code, L2_code, script_lookup)]
toefl[, geographic_dist := pair_distance_vec(L1_code, L2_code, geo_lookup)]
toefl[, genetic_dist := pair_distance_vec(L1_code, L2_code, gen_lookup)]
toefl[, syntactic_dist := pair_distance_vec(L1_code, L2_code, syn_lookup)]
toefl[, morphological_dist := pair_distance_vec(L1_code, L2_code, morph_lookup)]
toefl[, phonological_dist := pair_distance_vec(L1_code, L2_code, phon_lookup)]
toefl[, inventory_dist := pair_distance_vec(L1_code, L2_code, inv_lookup)]

```


```{r}

stex[, .(
  n = .N,
  missing_feat = sum(is.na(featural_dist)),
  missing_script = sum(is.na(script_dist)),
  missing_geo = sum(is.na(geographic_dist)),
  missing_gen = sum(is.na(genetic_dist))
)]

toefl[, .(
  n = .N,
  missing_feat = sum(is.na(featural_dist)),
  missing_script = sum(is.na(script_dist)),
  missing_geo = sum(is.na(geographic_dist)),
  missing_gen = sum(is.na(genetic_dist))
)]

```

```{r}

fwrite(stex, "data/stex_with_uriel_distances.csv")
fwrite(toefl, "data/toefl_with_uriel_distances.csv")

```


TODO:
- Think about what to do with Pashto
- Add stuff from [here](https://databank.worldbank.org/source/world-development-indicators/), maybe IT.NET.USER.ZS, NY.GDP.PCAP.PP.KD, SE.TER.ENRR, SE.XPD.TOTL.GD.ZS, SE.ADT.LITR.ZS, SP.URB.TOTL.IN.ZS, SM.POP.TOTL.ZS
- Add stuff from [here](https://hdr.undp.org/data-center/human-development-index#/indicies/HDI)
- Probably do the above only for stex
- For toefl and stex, download [ASJP](https://asjp.clld.org/) so we can use their distances as a competitor
- I thought about including walks/phoible based distances, but could get away with not, since they are included in uriel


# Extra Preprocessing

```{r}

stex.augmented <- read.csv("data/stex_with_uriel_distances.csv")
setDT(stex.augmented)

```


```{r}

dist_cols <- grep("(_dist$)|(^hyperbolic_dist$)", names(stex.augmented), value = TRUE)

setnames(
  stex.augmented,
  old = dist_cols,
  new = paste0(dist_cols, "_l1l2")
)

setnames(
  stex.augmented,
  old = c("morph", "lex", "new_feat", "new_sounds"),
  new = c("Morph_Score", "Lex_Score", "New_Feature_Score", "New_Sound_Score")
)

```


```{r}

dutch_code <- "dutc1256"

l1_dutch <- unique(
  stex.augmented[, .(L1_code)]
)

l1_dutch[, featural_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, feat_lookup)]
l1_dutch[, script_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, script_lookup)]
l1_dutch[, geographic_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, geo_lookup)]
l1_dutch[, genetic_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, gen_lookup)]
l1_dutch[, syntactic_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, syn_lookup)]
l1_dutch[, morphological_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, morph_lookup)]
l1_dutch[, phonological_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, phon_lookup)]
l1_dutch[, inventory_dist_l1dutch := pair_distance_vec(L1_code, dutch_code, inv_lookup)]

l2_dutch <- unique(
  stex.augmented[, .(L2_code)]
)

l2_dutch[, featural_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, feat_lookup)]
l2_dutch[, script_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, script_lookup)]
l2_dutch[, geographic_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, geo_lookup)]
l2_dutch[, genetic_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, gen_lookup)]
l2_dutch[, syntactic_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, syn_lookup)]
l2_dutch[, morphological_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, morph_lookup)]
l2_dutch[, phonological_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, phon_lookup)]
l2_dutch[, inventory_dist_l2dutch := pair_distance_vec(L2_code, dutch_code, inv_lookup)]

stex.augmented[l1_dutch, on = .(L1_code), `:=`(
  featural_dist_l1dutch = i.featural_dist_l1dutch,
  script_dist_l1dutch = i.script_dist_l1dutch,
  geographic_dist_l1dutch = i.geographic_dist_l1dutch,
  genetic_dist_l1dutch = i.genetic_dist_l1dutch,
  syntactic_dist_l1dutch = i.syntactic_dist_l1dutch,
  morphological_dist_l1dutch = i.morphological_dist_l1dutch,
  phonological_dist_l1dutch = i.phonological_dist_l1dutch,
  inventory_dist_l1dutch = i.inventory_dist_l1dutch
)]

stex.augmented[l2_dutch, on = .(L2_code), `:=`(
  featural_dist_l2dutch = i.featural_dist_l2dutch,
  script_dist_l2dutch = i.script_dist_l2dutch,
  geographic_dist_l2dutch = i.geographic_dist_l2dutch,
  genetic_dist_l2dutch = i.genetic_dist_l2dutch,
  syntactic_dist_l2dutch = i.syntactic_dist_l2dutch,
  morphological_dist_l2dutch = i.morphological_dist_l2dutch,
  phonological_dist_l2dutch = i.phonological_dist_l2dutch,
  inventory_dist_l2dutch = i.inventory_dist_l2dutch
)]

```


```{r}
fwrite(stex.augmented, "data/stex_with_uriel_distances_edit.csv")

```


```{r}

stex.new <- read.csv("data/stex_with_uriel_distances.csv")
setDT(stex.new)

```


```{r}

cols_to_suffix <- c(
  "island_dist",
  "speaker_dist",
  "hyperbolic_dist",
  "l2v_genetic_dist",
  "l2v_geographic_dist",
  "l2v_syntactic_dist",
  "l2v_inventory_dist",
  "l2v_phonological_dist",
  "l2v_featural_dist",
  "asjp_dist"
)

new_names <- paste0(cols_to_suffix, "_l1dutch")
data.table::setnames(stex.new, old = cols_to_suffix, new = new_names)

stex.augmented[, (intersect(names(stex.augmented), new_names)) := NULL]
stex.augmented[, (new_names) := stex.new[, ..new_names]]

```


```{r}

fwrite(stex.augmented, "data/stex_with_uriel_distances.csv")

```


# Add in other WDI/HDI covariates

```{r}

toefl.augmented <- read.csv("data/toefl_with_uriel_distances.csv")
stex.augmented <- read.csv("data/stex_with_uriel_distances.csv")
setDT(toefl.augmented)
setDT(stex.augmented)

```


```{r}

hdi <- read.csv("data/human-development-index/human-development-index.csv")
setDT(hdi)
wdi <- read.csv("data/wdi/wdi_data.csv")
setDT(wdi)

```

```{r}

country_map <- data.table(C = unique(stex.augmented$C))

country_map[, C_pretty := {
  y <- as.character(C)
  y <- gsub("([a-z])([A-Z])", "\\1 \\2", y, perl = TRUE)
  y <- gsub(",", ", ", y)
  y <- gsub("\\s+", " ", y)
  trimws(y)
}]

manual_iso <- c(
  "Congo,Republic" = "COG",
  "Congo,DemocraticRepublicofthe" = "COD",
  "Coted'Ivoire" = "CIV",
  "Korea,South" = "KOR",
  "CzechRepublic" = "CZE",
  "Turkey" = "TUR",
  "Syria" = "SYR",
  "Iran" = "IRN",
  "Egypt" = "EGY",
  "Russia" = "RUS",
  "Vietnam" = "VNM",
  "Venezuela" = "VEN",
  "Slovakia" = "SVK",
  "HongKong" = "HKG",
  "CapeVerde" = "CPV",
  "BosniaandHerzegovina" = "BIH",
  "SerbiaandMontenegro" = NA_character_,
  "Yugoslavia" = NA_character_,
  "USSR" = NA_character_,
  "Czechslowakia" = NA_character_
)

country_map[, iso3 := countrycode(
  C_pretty,
  origin = "country.name",
  destination = "iso3c",
  warn = FALSE
)]


```


```{r}

# most recent as of the pub date of stex data, which is 2019
# fallback is nearest post-2019
as_of_year <- 2019L

hdi_latest <- hdi[
  !is.na(Code) &
    Code != "" &
    !is.na(Human.Development.Index),
  {
    pre_idx <- which(Year <= as_of_year)

    if (length(pre_idx) > 0L) {
      .SD[pre_idx[which.max(Year[pre_idx])]]
    } else {
      post_idx <- which(Year > as_of_year)
      .SD[post_idx[which.min(Year[post_idx])]]
    }
  },
  by = Code
]

hdi_latest <- hdi_latest[
  ,
  .(
    iso3 = Code,
    hdi_entity = Entity,
    hdi_year = Year,
    hdi_timing = fifelse(
      Year <= as_of_year,
      "latest_available_at_or_before_2019",
      "fallback_closest_after_2019"
    ),
    hdi = Human.Development.Index,
    hdi_region = World.region.according.to.OWID
  )
]

```


```{r}

year_cols <- grep("^X[0-9]{4}", names(wdi), value = TRUE)

wdi_clean <- wdi[
  !is.na(Country.Code) &
    Country.Code != "" &
    !is.na(Series.Name) &
    Series.Name != "" &
    nchar(Country.Code) == 3
]

wdi_long <- melt(
  wdi_clean,
  id.vars = c("Country.Name", "Country.Code", "Series.Name", "Series.Code"),
  measure.vars = year_cols,
  variable.name = "year_col",
  value.name = "value_raw",
  variable.factor = FALSE
)

wdi_long[, year := as.integer(sub("^X([0-9]{4}).*$", "\\1", year_col))]

wdi_long[
  value_raw %chin% c("..", "", NA_character_),
  value_raw := NA_character_
]

wdi_long[, value := suppressWarnings(as.numeric(value_raw))]

as_of_year <- 2019L

wdi_latest <- wdi_long[
  !is.na(value),
  {
    pre_idx <- which(year <= as_of_year)

    if (length(pre_idx) > 0L) {
      .SD[pre_idx[which.max(year[pre_idx])]]
    } else {
      post_idx <- which(year > as_of_year)
      .SD[post_idx[which.min(year[post_idx])]]
    }
  },
  by = .(Country.Code, Country.Name, Series.Name, Series.Code)
]

wdi_latest[
  ,
  timing := fifelse(
    year <= as_of_year,
    "latest_available_at_or_before_2019",
    "fallback_closest_after_2019"
  )
]

```


```{r}

wdi_name_map <- c(
  "Individuals using the Internet (% of population)" =
    "wdi_internet_pct",

  "GDP per capita, PPP (constant 2021 international $)" =
    "wdi_gdp_pc_ppp_const2021",

  "School enrollment, tertiary (% gross)" =
    "wdi_tertiary_enrollment_pct",

  "School enrollment, tertiary, female (% gross)" =
    "wdi_tertiary_enrollment_female_pct",

  "School enrollment, tertiary, male (% gross)" =
    "wdi_tertiary_enrollment_male_pct",

  "Government expenditure on education, total (% of GDP)" =
    "wdi_gov_education_expenditure_pct_gdp",

  "Literacy rate, adult total (% of people ages 15 and above)" =
    "wdi_adult_literacy_pct",

  "Urban population (% of total population)" =
    "wdi_urban_population_pct"
)

wdi_latest[, feature := unname(wdi_name_map[Series.Name])]

wdi_latest[
  is.na(feature),
  feature := paste0("wdi_", make.names(Series.Code))
]

wdi_latest[, feature_year := paste0(feature, "_year")]
wdi_latest[, feature_timing := paste0(feature, "_timing")]
```


```{r}

wdi_values_wide <- dcast(
  wdi_latest,
  Country.Code + Country.Name ~ feature,
  value.var = "value"
)

wdi_years_wide <- dcast(
  wdi_latest,
  Country.Code + Country.Name ~ feature_year,
  value.var = "year"
)

wdi_timing_wide <- dcast(
  wdi_latest,
  Country.Code + Country.Name ~ feature_timing,
  value.var = "timing"
)

wdi_country_features <- Reduce(
  function(x, y) merge(x, y, by = c("Country.Code", "Country.Name"), all = TRUE),
  list(wdi_values_wide, wdi_years_wide, wdi_timing_wide)
)

setnames(
  wdi_country_features,
  old = c("Country.Code", "Country.Name"),
  new = c("iso3", "wdi_country_name")
)

```


```{r}


country_features <- merge(
  country_map,
  hdi_latest,
  by = "iso3",
  all.x = TRUE
)

country_features <- merge(
  country_features,
  wdi_country_features,
  by = "iso3",
  all.x = TRUE
)

```


```{r}

stex_augmented_hdi_wdi <- copy(stex.augmented)
stex_augmented_hdi_wdi[, row_id__ := .I]

stex_augmented_hdi_wdi <- merge(
  stex_augmented_hdi_wdi,
  country_features,
  by = "C",
  all.x = TRUE,
  sort = FALSE
)

setorder(stex_augmented_hdi_wdi, row_id__)
stex_augmented_hdi_wdi[, row_id__ := NULL]

stopifnot(nrow(stex_augmented_hdi_wdi) == nrow(stex.augmented))

```


```{r}

fwrite(stex_augmented_hdi_wdi, "data/stex_with_indicators.csv")

```




